Eagle: Making multiple-locus association mapping on a genome-wide scale routine

### 2018-07-10 AM+draft.tex
3Jul revised table sup.

\begin{abstract}
Since the earliest of genome-wide association studies, a key shortcoming in how their data have been analysed has persisted. The strength of association between a marker locus and trait is measured for each locus separately, on a locus-by-locus basis. Multiple-locus methods that map multiple-locus trait associations simultaneously have been available for some time. However, they have attracted little attention. They can be demanding, computationally, and their results are not always easy to interpret. Yet, it is widely accepted that multiple-locus methods are superior, statistically, to locus-by-locus methods. Here, we present our method that makes the multiple-locus analysis of data from genome wide association studies routine. It is formulated within a linear mixed model framework. 
We call our method AMplus.  AMplus produces results faster than competing multiple-locus methods and often with greater statistical power. Also, it is just a little slower than the fastest single-locus linear mixed model implementations. 
AMplus is freely available as a fully documented R package. 
\end{abstract}

Over the past decade,  Genome wide association studies (GWAS) have changed considerably in both their analysis and design. Early GWAS
 followed a case-control design. Association mapping methods were no more complicated than contingency table tests or simple 
linear regression. These designs though had a tendency to yield spurious findings if there was unrecognised population stratification. This prompted a shift towards family-based designs and score tests, such as the tdt test and its variants (refs). Today, instead of by design, it is through statistical modelling that we account for the effects of population stratification. This has meant that data can be collected from general populations, even if these populations are highly structured. Analysis via sophisticated association mapping methods, such as linear mixed model based approaches,  is now almost routine.

What has not changed over the past decade is that it remains common practice to analyse gwas data on a locus by locus basis. This is despite there being several significant problems with analysing data in this way. First, the aim of association mapping is to identify regions of the genome that house genes that are influencing a trait. The identification of these regions from these analyses is not always straightforward. GWAS results are reported, typically, via Manhattan plots that plot the -log10 of the p value for each locus against the map position of the locus. The location of peaks in this plot signify regions of interest. Inferring the exact number of regions of interest can be difficult If the peaks are not well separated. Second, when multiple statistical tests are performed, the probability of wrongly accepting a significant result (type 1 error) is inflated. This is known as the multiple testing problem (refs). Many different solutions have been offered (refs). Yet, there is still no well accepted way of correcting for multiple testing in the context of genome- wide association mapping. Third, many of the traits whose genetic secretes we are trying to discover are complex. There will be multiple loci in linkage equilibrium with genes that influence the trait. Yet, A locus by locus mapping approach only assesses the evidence for association between a single marker locus and trait.

It is somewhat surprising then that multiple-locus association mapping methods haven't attracted more attention. Methods based on 
regularisation techniques, such as ridge regression and lasso, measure all locus-trait associations simultaneously. 
Here, multiple testing is not an issue. These techniques though are computationally demanding. Also, their results can be difficult to interpret. The strength of association is not measured by a p-value but by the size of the regression coefficient for a locus in the model. More recently, associations have started to be mapped with random forests (refs). Similar to regularisation techniques though, it is not clear how to infer genomic regions of interest from their findings (refs). An multiple-locus method that does show promise is the multi-locus linear mixed model method (ref). The best multiple-locus model is built with simple variable selection. Results are immediately interpretable but here, computation can be a challenge for large data.

In this paper, we present our new multiple-locus method for genome wide association mapping, which we are calling Eagle. Eagle combines the strength of regularisation techniques (being able to fit all locus-trait associations jointly), with simple variable selection (having easy to interpret results). Our method does not require a significance threshold to be set nor regularisation parameters to be fine tuned. Through a clever dimension reduction step, we are able to achieve a computational performance similar to the fastest single-locus linear mixed model implementations. Eagle is statistically more powerful than single-locus association mapping and is as and often more powerful than 
most multiple-locus methods. Our aim is to make multiple-locus association mapping on a genome wide scale routine. To this end, we have created a fully documented R package, that is easy to use, even for non R users.  Our package accepts marker data of different 
formats and can handle data larger than a computer's  memory capacity. It includes detailed error checking and makes heavy use of 
distributed computing for computation when available.  The purpose of this work was to make multiple-locus association mapping on 
a genome-wide scale, for large data sets,  practical and we have built AMplus accordingly.

@@

 We list some of the key features that make computer programs for association mapping useful in 
Supplementary Table \ref{suptabsummary}. We also identify which of these 
features are present in Eagle and the other seven implementations. For example, all but glmnet were purpose-built for 
association mapping. Not all the computer programs come with a user manual.  Most could accomodate data on additional fixed effects but only Eagle and FaST-LMM could deal with data larger than the memory capacity of the computer. Also, if 
the input data contained errors, we found only Eagle and FaST-LMM tried to diagnose the problem and 
provided advice on how to fix the errors.

>>>>

\begin{abstract}

For over a decade, methods for genome-wide association mapping have become progressively more sophisticated, statistically and computationally. Yet a well know limitation has persisted, the strength of association between a locus and trait is assessed on a locus-by-locus basis. Superior multiple-locus methods that detect multiple SNP-trait associations, simultaneously, have been available for some time but their application has been minimal. Here, we present Eagle for multiple-locus association mapping. Our goal was to develop a method that shifts mainstream association mapping from being single-locus to multiple-locus. We will show through the analysis of simulated and mouse data that Eagle is more powerful than competing single- and multiple-locus methods and can be orders of magnitude faster. Eagle has been implemented as an R package, with an easy to use web-based Graphical User Interface (GUI) for users unfamiliar with R.
\end{abstract}

Over the past decade,  genome-wide association studies have changed considerably in both their analysis and design. Early studies
 followed a case-control design. Association mapping methods were no more complicated than contingency table tests or simple 
linear regression. These designs though had a tendency to yield spurious findings if there was unrecognised population stratification. This prompted a shift towards family-based designs and score tests, such as the tdt test and its variants (refs). Today, instead of by design, it is through statistical modelling that we account for the effects of population stratification. This has meant that data can be collected from general populations, even if these populations are highly structured. Analysis via sophisticated association mapping methods, such as linear mixed model based approaches,  is now almost routine.

What has not changed over the past decade is that it remains common practice to analyse genome-wide association study (GWAS) data on a locus-by-locus basis. This is despite there being several significant problems with analysing data in this way. First, the aim of association mapping is to identify regions of the genome that house genes that are influencing a trait. The identification of these regions from these analyses is not always straightforward. GWAS results are reported, typically, via Manhattan plots that plot the -log$_{10}$ of the P value for each locus against the map position of the locus. The location of peaks in this plot signify regions of interest. Inferring the exact number of regions of interest can be difficult if the peaks are not well separated. Second, when multiple statistical tests are performed, the probability of wrongly accepting a significant result (type 1 error) is inflated. This is known as the multiple testing problem (refs). Many different solutions have been offered (refs). Yet, there is still no well accepted way of correcting for multiple testing in the context of genome-wide association mapping. Third, many of the traits whose genetic secretes we are trying to discover are complex. There will be multiple loci in linkage equilibrium with genes that influence the trait. Yet, a locus-by-locus mapping approach only assesses the evidence for association between a single marker locus and trait.

It is somewhat surprising then that multiple-locus association mapping methods haven't attracted more attention. Methods based on 
regularisation techniques, such as ridge regression and lasso, measure all locus-trait associations simultaneously. 
Here, multiple testing is not an issue. These techniques though are computationally demanding. Also, their results can be difficult to interpret. The strength of association is not measured by a P value but by the size of the regression coefficient for a locus in the model. More recently, associations have started to be mapped with random forests (refs). Similar to regularisation techniques though, it is not clear how to infer genomic regions of interest from their findings (refs). A multiple-locus method that does show promise is the multiple-locus linear mixed model method (ref). The best multiple-locus model is built with simple variable selection. Results are immediately interpretable but here, computation can be a challenge if the data set is large.

In this paper, we present our new multiple-locus method for genome wide association mapping, which we are calling Eagle. Eagle combines the strength of regularisation techniques (being able to fit all locus-trait associations jointly), with simple variable selection (having easy to interpret results). Our method does not require a significance threshold to be set. Through a simple dimension reduction step, we are able to achieve a computational performance similar to the fastest single-locus linear mixed model implementations. 
Our aim was to make multiple-locus association mapping on a genome wide scale routine. To this end, we have implemented Eagle 
within an R package of the same name. 
Our package accepts marker data of different 
formats,  can handle data larger than a computer's  memory capacity, and makes heavy use of 
distributed computing for computation when available.

@@

 We list some of the key features that make computer programs for association mapping useful in 
Supplementary Table \ref{suptabsummary}. We also identify which of these 
features are present in Eagle and the other seven implementations. For example, all but glmnet were purpose-built for 
association mapping. Not all the computer programs come with a user manual.  Most could accomodate data on additional fixed effects but only Eagle and FaST-LMM could deal with data larger than the memory capacity of the computer. Also, if 
the input data contained errors, we found only Eagle and FaST-LMM tried to diagnose the problem and 
provided guidance on how to fix the errors.

### 2018-07-10 supmaterials.tex
3Jul revised table sup.

\documentclass{article}
\usepackage{graphicx}
\usepackage{authblk}
\usepackage{amsfonts}
\usepackage{pifont}
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%

@@

\title{Eagle: Making multiple-locus association mapping on a genome-wide scale routine}
\author[1]{Andrew W. George}
\author[1]{Arunas Verbyla}
\author[2]{Joshua Bowden}
\author[1]{Some other authors}

@@

\begin{table}
\label{suptabsummary}
\caption{A summary of the features possessed by the Eagle package and the comparison implementations.}
\begin{tabular}{lccccccccc}
                                &  \multicolumn{9}{c}{Computer Software for Association Mapping} \\ \cline{2-10}
                                 & \multicolumn{6}{c}{Multiple-locus}  & & \multicolumn{2}{c}{Single-locus} \\  \cline{2-7}   \cline{9-10}
 Features                 & Eagle  & bigRR & glmnet & LMM-Lasso & MLMM & r2VIM     & & FaST-LMM & GEMMA \\  \hline
Purpose built$^1$    &   \cmark     &    \cmark      &  \xmark   &  \cmark  &  \cmark  &  \cmark && \cmark & \cmark           \\ [0.2cm]
Well documented$^2$           & \cmark &   \xmark   & \cmark   & \xmark     &   \cmark     & \xmark          && \cmark    &    \cmark    \\  [0.2cm]
Simultaneous   &    \cmark     &    \cmark        &   \cmark          &      \cmark             &  \xmark          &  \xmark              &&   \xmark   &  \xmark     \\  
fitting of SNPs     &         &            &             &                   &            &                &&      &      \\  [0.2cm]
Additional fixed effects$^3$          &   \cmark   &      \cmark      &     \cmark    &  \xmark                 &      \cmark      &      \cmark          &&      \cmark & \cmark     \\   [0.2cm]
Data larger than RAM$^4$                   &   \cmark    &      \xmark      & \xmark          &  \xmark    &  \xmark    &   \xmark     && \cmark      & \xmark    \\  [0.2cm]
Threshold free$^5$        &    \cmark     &  \xmark    &  \xmark         &    \xmark               &     \cmark       &  \xmark              &&  \xmark   & \xmark      \\  [0.2cm]
Informative error checking           &  \cmark      &     \xmark        &      \xmark     &           \xmark        &      \xmark       &         \xmark       && \cmark  & \xmark \\ \hline

\end{tabular}
{$^1$ \scriptsize{Computer software specifically designed for the analysis of data from GWAS.}}\\
{$^2$ \scriptsize{More than just a readme file or comments in an example file. Those programs with ticks had detailed user manuals.}}\\
{$^3$ \scriptsize{Ability to accommodate additional fixed effects in the model such as age, sex, and population structure effects.}} \\
{$^4$ \scriptsize{Able to deal with data larger than the memory capacity of the computer.}} \\
{$^5$ \scriptsize{Results reported as the set of SNP closest to the genes influencing a trait. No need to construct thresholds to determine 
significance of the findings.}}\\
{$^6$  \scriptsize{All the programs terminated on errors. However, not all the programs informed the user of the cause and how to fix the errors.}}\\
\end{table}

\begin{figure}
\caption{Memory usage (in gigabytes) of Eagle and the other association mapping programs across 
the six simulation scenarios. The maximum amount of memory on the computer is 128 gigabytes. 
The x-axis is on the log scale. GEMMA, a single-locus implementation, had the lowest memory usage. 
Of the multiple-locus implementations, Eagle had the lowest memory usage. Also, it 
was the only multiple-locus 
implementation able to produce results for data under  scenario 10000 x 1.5M. This is due to its ability 
to handle data larger than the available memory of a computer. FaST-LMM was run where all the SNP data are used 
to estimate the relationship matrix (FaST-LMM$^{all}$)   and where genotype data from every five-hundredth SNP are used to 
estimate the relationship matrix (FaST-LMM$^{few}$)}.
\begin{center}
\includegraphics[width=15cm, height=15cm]{mem.eps}
\end{center}
\end{figure}

>>>>

\documentclass{article}
\usepackage{xr}
\usepackage{lscape}

\usepackage{graphicx}
\usepackage{authblk}
\usepackage{amsfonts}
\usepackage{pifont}

\renewcommand{\tablename}{Supplementary Table}
\renewcommand{\figurename}{Supplementary Figure}
\newcommand{\rowgroup}[1]{\hspace{-1em}#1}

@@

\title{Eagle: Making multiple-locus association mapping on a genome-wide scale routine}
\author[1]{Andrew W. George}
\author[1]{Arunas Verbyla}
\author[2]{Joshua Bowden}

@@

\begin{landscape}

\begin{table}
\caption{{\bf Implementation and methodology attributes of eight computer programs/packages for genome-wide association mapping.} For the different types of model, 
LMM is linear mixed model.  GLM is generalised linear model., GLMM is generalised linear mixed model, and RF is random forests. 
}
\label{suptabsummary}
\vspace{0.5cm}
\begin{tabular}{lcccccccc} \hline
                                                   Attributes                  & {\bf Eagle}                                 & {\bf bigRR}             & {\bf glmnet}            & {\bf LMM-Lasso}                    & {\bf MLMM} & {\bf r2VIM}      & {\bf FaST-LMM} & {\bf GEMMA} \\  \hline
{\em Implementation}    &         &            &             &                   &            &                &      &      \\ 
\hspace{1mm}  Purpose built$^a$    &   Yes     &    Yes      &  No   &   Yes  &  Yes  &  Yes  & Yes  & Yes          \\ [0.35cm]

\hspace{1mm}  Language                 &  R/C++       &    R        &      R       &     Python     &  R          &    R         &  C++ and     &   C++   \\  
                                                          &         &            &             &                   &            &                                     &   Python$^b$       &      \\  [0.35cm]

\hspace{1mm} GUI                            & Yes &    No      & No          &  No    &  No    &   No     & No     & No    \\  [0.35cm]

\hspace{1mm} Documentation        &  Videos,                &    R help       &  Vignettes,    &   Readme.txt,       &   Vignette,          &     R help           &   Videos,              &  User-manual,     \\  
                                                        &   user-manuals,    &                      &  R help         &    test script           &  R help              &                           &    user-manuals,  &   website  \\
                                                        &   website,    R help          &                      &                     &                              &                           &                           &    website            &    \\   [0.35cm]

\hspace{1mm} Additional fixed        &   Yes   &      Yes      &     Yes    &  No                 &      Yes      &      Yes          &      Yes & Yes     \\  
\hspace{1mm} effects$^c$               &           &                   &                 &                   &                   &                     &              &   adsf    \\  [0.35cm]

\hspace{1mm} Types of  &  Cont.      &   Cont.,          &   Cont.,           &   Cont.                 &   Cont.          &     Cont.           &   Cont.   &  Cont.    \\  
\hspace{1mm} trait data  &                &    binary,      &      binary,         &                   &            &                &      &      \\  
                                      &                &   count         &       count          &                   &            &                &      &      \\  [0.35cm]

\hspace{1mm} Data larger                   &   Yes    &      No      & No          &  No    &  No    &   No     & Yes      & No    \\  
\hspace{1mm} than memory                                             &         &            &             &                   &            &                &      &      \\  [0.8cm]

{\em Methodology}    &         &            &             &                   &            &                &      &      \\  
\hspace{1mm}  Model                     & LMM   &   &  GLMM &   GLM  & LMM  & RF & LMM   & LMM \\ [0.15cm]

\hspace{1mm} SNPs fitted       &    All/multiple      &    All       &   All        &      All          &  Multiple          &  Multiple              &   Single   &  Single   \\  [0.35cm]

\hspace{1mm}  Selection type              & Model   & Variable  &  Variable &  Variable  &  Model  &  Variable & Variable  & Variable \\  [0.35cm]

\hspace{1mm} Threshold free$^5$        &    Yes     &  No    &  No         &    No               &     Yes       &  No              &  No   & No      \\  [0.15cm]  \hline

\end{tabular}
{$^a$ \scriptsize{Specifically created for the analysis of GWAS data.}}\\
{$^b$ \scriptsize{Separate programs, one written in Python, the other C++}} \\
{$^c$ \scriptsize{Capacity for additional fixed effects (such as  age, sex, and/or population structure effects) to be included directly in the model.}} \\
\end{table}

\end{landscape}

\begin{landscape}

\begin{table}
\caption{Some of the key features possessed by Eagle and the other seven computer programs/packages for association mapping}
\label{suptabsummary}

\begin{tabular}{lccccccccc} 
                                &  \multicolumn{9}{c}{Computer Programs/Packages for Association Mapping} \\ \cline{2-10}
                                 & \multicolumn{6}{c}{Multiple-locus}  & & \multicolumn{2}{c}{Single-locus} \\  \cline{2-7}   \cline{9-10}
 Features                 & Eagle  & bigRR & glmnet & LMM-Lasso & MLMM & r2VIM     & & FaST-LMM & GEMMA \\  \hline
Purpose built$^1$    &   Yes     &    Yes      &  No   &  Yes  &  Yes  &  Yes && Yes & Yes           \\ [0.2cm]
Well documented$^2$           & Yes &   No   & Yes   & No     &   Yes     & No          && Yes    &    Yes    \\  [0.2cm]
Simultaneous   &    Yes     &    Yes        &   Yes          &      Yes             &  No          &  No              &&   No   &  No     \\  
fitting of SNPs     &         &            &             &                   &            &                &&      &      \\  [0.2cm]
Additional fixed effects$^3$          &   Yes   &      Yes      &     Yes    &  No                 &      Yes      &      Yes          &&      Yes & Yes     \\   [0.2cm]
Data larger than RAM$^4$                   &   Yes    &      No      & No          &  No    &  No    &   No     && Yes      & No    \\  [0.2cm]
Threshold free$^5$        &    Yes     &  No    &  No         &    No               &     Yes       &  No              &&  No   & No      \\  [0.2cm]
Informative error checking           &  Yes      &     No        &      No     &           No        &      No       &         No       && Yes  & No \\ \hline

\end{tabular}
{$^1$ \scriptsize{Specifically created for the analysis of GWAS data.}}\\
{$^2$ \scriptsize{More than just a readme file or comments in an example file. Programs/packages with a tick had a detailed user manual.}}\\
{$^3$ \scriptsize{Ability to handle additional fixed effects such as  age, sex, and/or population structure effects.}} \\
{$^4$ \scriptsize{Able to deal with data larger than the memory capacity of the computer.}} \\
{$^5$ \scriptsize{Results reported as the set of SNP closest to the genes influencing a trait. No need to construct thresholds to determine 
significance of the findings.}}\\
{$^6$  \scriptsize{All the programs terminated on errors. However, not all the programs informed the user of the cause and how to fix the errors.}}\\
\end{table}

\end{landscape}

\begin{figure}
\caption{Memory usage (in gigabytes) of Eagle and the other association mapping programs across 
the six simulation scenarios. The maximum amount of memory on the computer is 128 gigabytes. 
The x-axis is on the log scale. GEMMA, a single-locus implementation, had the lowest memory usage. 
Of the multiple-locus implementations, Eagle had the lowest memory usage. Also, it 
was the only multiple-locus 
implementation able to produce results for data under  scenario 10000 x 1.5M. This is due to its ability 
to handle data larger than the available memory of a computer. FaST-LMM was run where all the SNP data are used 
to estimate the relationship matrix (FaST-LMM$^{all}$)   and where genotype data from every five-hundredth SNP are used to 
estimate the relationship matrix (FaST-LMM$^{few}$)}.
\label{supfigmem}

\begin{center}
\includegraphics[width=15cm, height=15cm]{mem.eps}
\end{center}
\end{figure}

\begin{figure}
\caption{Power verse false discovery rates for Eagle and the single-locus methods GEMMA and FaST-LMM. 
FaST-LMM was run where all the SNP data are used 
to estimate the relationship matrix (FaST-LMM$^{all}$)   and where genotype data from every five-hundredth SNP are used to 
estimate the relationship matrix (FaST-LMM$^{few}$). Eagle has substantially higher power than the single-locus methods.   }

\label{supfigpowersingle}
\begin{center}
\includegraphics[width=15cm, height=15cm]{power2main.eps}
\end{center}

\end{figure}
